WSClean: an implementation of a fast, generic wide-field imager

### 2014-01-29 articles/2014/Article-WSClean/wsclean.tex
Progress

\documentclass[useAMS,usenatbib]{mn2e}
\usepackage[utf8]{inputenc}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
%\usepackage[font={it},labelfont={bf}]{caption}
%\usepackage{units}
\usepackage{times}
\usepackage{color}
\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}

@@

%\newcommand{\editmark}[1]{#1}
%\newcommand{\editmark}[1]{{\color{red}{\textbf{#1}}}}

@@

\title[WSClean: a new imager]{WSClean: a new implementation of a fast, generic wide-field imager}

@@

To derive the w-stacking technique, Eq.~\eqref{eq:visibility-function} is rewritten to
\begin{align}\notag
V(u,v,w) = & \iint \frac{I'(l,m) e^{-2\pi i w(\sqrt{1-l^2-m^2}-1)}}{\sqrt{1-l^2-m^2}} \cdot \\ \notag
& e^{-2\pi i \left(ul + vm\right)} dl dm.
\end{align}
This is an ordinary two-dimensional Fourier transform going from $u,v$ space to $l,m$ space, and can be inverted to get:
% Intermediate step:
%\begin{align}\notag
%\frac{I'(l,m) e^{-2\pi i w(\sqrt{1-l^2-m^2}-1)}}{\sqrt{1-l^2-m^2}} = & \iint V(u,v,w) \cdot \\ \notag
%& e^{-2\pi i \left(ul + vm\right)} du dv,
%\end{align}
%which can be rewritten to
\begin{align}\notag
\frac{I'(l,m)}{\sqrt{1-l^2-m^2}} = & e^{2\pi i w(\sqrt{1-l^2-m^2}-1)} \iint V(u,v,w) \cdot \\ \notag
& e^{2\pi i \left(ul + vm\right)} du dv.
\end{align}
Integrating both sides over $w_0$ to $w_e$, the minimum and maximum value of $w$, results in
\begin{align}\notag
\frac{I'(l,m)\left(w_e - w_0\right)}{\sqrt{1-l^2-m^2}} = \int\limits_{w_0}^{w_e} e^{2\pi i w(\sqrt{1-l^2-m^2}-1)} \cdot \\ \label{eq:wstacking}
\iint V(u,v,w)  e^{2\pi i \left(ul + vm\right)} du dv dw.
\end{align}
The final step is to make the $u,v,w$ parameters discrete, so that the integration over $u$ and $v$ can become an inverse FFT and the integration over $w$ becomes a summation. This shows that the sky function can be reconstructed by i)~taking all samples that correspond to a specific value of $w$; ii)~grid those samples on a uniform grid; iii)~calculating the inverse FFT; iv)~apply the direction dependent phaseshift $e^{2\pi i w(\sqrt{1-l^2-m^2}-1)}$; v)~repeat this for all $w$-values and add the results together; vi)~apply the final scaling.

@@

TODO: discuss visibility prediction? (inverse of imaging)

@@

The new imager, called ``WSClean'', is written in the C++ language. It reads visibilities from CASA measurement sets and writes output images to Flexible Image Transport System (FITS) files. Several steps are multi-threaded: reading and writing; gridding from and to different $w$-layers; performing the Fast Fourier transforms; and minor Clean iterations (peak-finding and image subtraction). For the latter, intrinsics are used as well. Because of these optimizations, minor Cleaning iterations are so fast that performing the Clark Clean optimization \citep{clark-clean}, which consists of considering only a subset of pixels with a reduced PSF, is not so relevant. However, the PSF varies for imaging with a non-zero w-value, and subtracting a constant PSF image space leads to inaccuracies. After a number of these minor iterations, it is therefore beneficial to invert the model back to the visibilities and subtract the model directly from the visibilities. This is similar to the Cotton-Schwab cleaning method \citep{cotton-schwab-clean}.

@@

\subsection{Full-sky imaging}
For low-frequency telescopes, it can be useful to do full-sky imaging. In the case of MWA, the tile beam can have strong sidelobes at a distance of more than 90 degrees from the pointing centre. Imaging these sidelobes might be relevant because it might either be scientifically interesting (e.g. when searching for transients), or be useful for self-calibration or full-sky deconvolution. For example, the latter is useful when a strong source in a sidelobe affects the sensitivity near the pointing centre.

To image the full sky, an observation is split into short snapshots, their phase centre is changed to zenith and subsequently a suitable area is imaged. Depending on desired resolution and dynamic range, this might require very large images. To make an image at the MWA resolution, the image needs to have approximately 10K pixels in width and height. The $w$-values will be very small at zenith, because only the differential heights of the antennas will contribute to $w$-value at zenith. For the MWA, the maximum height difference between tiles is $9.3$m. The MWA tiles are in fact on a slight slope, and by fitting a plane to the antennas and changing the phase centre towards the normal of that plane, the $w$-value decreases to $7.2\textrm{m} / \lambda$.

(TODO: these $w$ values are in 2 min obses; need shorter time interval for correct values).

\subsection{W-snapshots} \label{sec:snapshot-imaging}
\begin{figure*}
\begin{center}
\includegraphics[width=8cm]{img/vela-normal-projection}
\includegraphics[width=8cm]{img/vela-zenith-projection}
\caption{13-min MWA observation of supernova remnants Vela and Puppis A. Left: normal projection centred on Puppis A, right: phase-rotated to zenith to reduce $w$-values, shifted towards Puppis A during imaging. Both images have been made with WSClean using the Cotton-Schwab clean algorithm. No beam correction was applied. TODO time difference and more discussion, replace with deeper images.}
\label{fig:vela-projection-example}
\end{center}
\end{figure*}
To get optimal imaging performance, the $w$-values can be kept small by phase-changing towards zenith. As shown, all-sky imaging can be quite efficient with this technique. However, if one is not interested in imaging the whole sky, and the direction of interest is far from zenith, the computational overhead of making all-sky images is undesirable.

Instead of imaging the whole sky, the image can be recentred from $(l,m)$ to $(\hat{l},\hat{m})=(l+\Delta l,m+\Delta m)$ by performing the substitution $(l,m)\leftarrow(\hat{l},\hat{m})$ in Eq.~\eqref{eq:wstacking}:
%When the Fourier transform is performed, the $w$ term has been corrected (either by w-stacking or by w-projection). Therefore, if $U(u, v)$ is the input to the 2-dimensional Fourier transform, the transform is described by
\begin{align}\notag
\frac{I'(\hat{l}, \hat{m})\left(w_e - w_0\right)}{\sqrt{1-\hat{l}^2-\hat{m}^2}} = \int\limits_{w_0}^{w_e} e^{2\pi i w(\sqrt{1-\hat{l}^2-\hat{m}^2}-1)} \cdot \\
\iint e^{2 \pi i \left(u\Delta l + v\Delta m\right)} V(u,v,w) e^{2\pi i \left(ul + vm\right)} du dv dw.
\end{align}
In words, recentering an image involves accounting for the position shift during the $w$-correction and final scaling, and shifting the visibilities in phase by multiplication with $e^{2 \pi i \left(u\Delta l + v\Delta m\right)}$ prior to the imaging. By doing the inverse corrections during visibility prediction, a ``shifted'' visibility set can be cleaned with major iterations similar to a non-shifted set.

A recentred image is in the projection of the zenith tangent plane, and will need an additional regridding step before multiple snapshots can be added together. An example of the difference in projection is given in Fig.~\ref{fig:vela-projection-example}, which displays a 13~min MWA observation imaged with WSClean. Cleaning an image that is in a different projection yields slightly different results, but qualitativily the two images are of equal accuracy. A recentred image can be stored in a Flexible Image Transport System (FITS) file, which allows setting the centre of tangent projection with the \texttt{CRPIXi} keyword \citep{wcs-in-fits}. Common viewers such as Kvis and DS9 support such FITS files and display their coordinates correctly. TODO show difference.

@@

\subsection{Application of WSClean to the MWA}
The MWA correlator splits observations in data products of typically a few minutes
TODO

\section{Analysis} \label{sec:analysis}
TODO: Show image quality; show time-efficiency comparisons.

\subsection{Gridding} \label{sec:gridding}
During gridding of the visibilities in $uv$-space, it is useful to apply a gridding convolution kernel. A common kernel function is a windowed sinc function. A low-pass filter decreases the flux of sources outside the field of view, and thus helps to attenuate aliased ghost sources and sidelobes (see \citealt{post-correlation-filtering}) that arise because of the assumed periodicity during the inverse FFT. By supersampling, a convolution kernel also makes it possible to place samples more accurately at their $uv$ position, thereby lowering decorrelation.

@@

\begin{table}
 \caption{Parameter values during benchmarks (unless the effect of the parameter was tested).}
 \begin{tabular}{lr}
   \textbf{Parameter} & \textbf{Value} \\
   Image size             & 3072 \\
   Number of visibilities & $10^9$
 \end{tabular}
\end{table}

@@

\begin{figure}
\begin{center}
\includegraphics[width=8cm]{img/benchmark-zenith-angle/za}
\caption{...}
\label{fig:timing-zenith-angle}
\end{center}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[width=8cm]{img/benchmark-nsamples/nsamples}
\caption{...}
\label{fig:timing-nsamples}
\end{center}
\end{figure}

>>>>

\documentclass[useAMS,usenatbib]{mn2e}
\usepackage[utf8]{inputenc}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
%\usepackage[font={it},labelfont={bf}]{caption}
%\usepackage{units}
\usepackage{times}
\usepackage{color}
\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}
\usepackage{xspace}

@@

%\newcommand{\editmark}[1]{#1}
%\newcommand{\editmark}[1]{{\color{red}{\textbf{#1}}}}
\newcommand{\degree}{\ensuremath{^{\circ}}\xspace}

@@

\title[WSClean: a fast, generic wide-field imager]{WSClean: an implementation of a fast, generic wide-field imager}

@@

To derive the w-stacking technique, Eq.~\eqref{eq:visibility-function} is rewritten to
\begin{align}\notag
V(u,v,w) = & \iint \frac{I'(l,m) e^{-2\pi i w(\sqrt{1-l^2-m^2}-1)}}{\sqrt{1-l^2-m^2}} \cdot \\ \notag
& e^{-2\pi i \left(ul + vm\right)} dl dm.
\end{align}
This is an ordinary two-dimensional Fourier transform going from $u,v$ space to $l,m$ space, and can be inverted to get:
% Intermediate step:
%\begin{align}\notag
%\frac{I'(l,m) e^{-2\pi i w(\sqrt{1-l^2-m^2}-1)}}{\sqrt{1-l^2-m^2}} = & \iint V(u,v,w) \cdot \\ \notag
%& e^{-2\pi i \left(ul + vm\right)} du dv,
%\end{align}
%which can be rewritten to
\begin{align}\notag
\frac{I'(l,m)}{\sqrt{1-l^2-m^2}} = & e^{2\pi i w(\sqrt{1-l^2-m^2}-1)} \iint V(u,v,w) \cdot \\ \notag
& e^{2\pi i \left(ul + vm\right)} du dv.
\end{align}
Integrating both sides over $w_0$ to $w_e$, the minimum and maximum value of $w$, results in
\begin{align}\notag
\frac{I'(l,m)\left(w_e - w_0\right)}{\sqrt{1-l^2-m^2}} = \int\limits_{w_0}^{w_e} e^{2\pi i w(\sqrt{1-l^2-m^2}-1)} \cdot \\ \label{eq:wstacking}
\iint V(u,v,w)  e^{2\pi i \left(ul + vm\right)} du dv dw.
\end{align}
The final step is to make the $u,v,w$ parameters discrete, so that the integration over $u$ and $v$ can become an inverse FFT and the integration over $w$ becomes a summation. This shows that the sky function can be reconstructed by i)~grid samples with equal $w$-value on a uniform grid; ii)~calculating the inverse FFT; iii)~apply the direction dependent phaseshift $e^{2\pi i w(\sqrt{1-l^2-m^2}-1)}$; iv)~repeat this for all $w$-values and add the results together; v)~apply the final scaling.

@@

The inverse of imaging, i.e., to calculate the visibility from a model image, can be done by reversing the $w$-stacking algorithm: i)~inverse scale the image; ii)~copy the image to several layers; iii)~inverse apply the direction-dependent phaseshift for each layer; iv)~FFT each layer; and v)~sample a required visibility from the correct $w$-layer. We will refer to this operation as prediction.

@@

The new imager, called ``WSClean'', is written in the C++ language. It reads visibilities from CASA measurement sets and writes output images to Flexible Image Transport System (FITS) files. Several steps are multi-threaded: reading and writing; gridding from and to different $w$-layers; performing the Fast Fourier transforms; and minor Clean iterations (peak-finding and image subtraction). For the latter, intrinsics are used as well. Because of these optimizations, minor Cleaning iterations are so fast that performing the Clark Clean optimization \citep{clark-clean}, which consists of considering only a subset of pixels with a reduced PSF, is not so relevant. However, the PSF varies for imaging with a non-zero w-value, and subtracting a constant PSF in image space leads to inaccuracies. After a number of these minor iterations, it is therefore beneficial to invert the model back to the visibilities and subtract the model directly from the visibilities \citep{wprojection-cornwell}. This is similar to the Cotton-Schwab cleaning method \citep{cotton-schwab-clean}.

@@

\subsection{Full-sky imaging}
For low-frequency telescopes, it can be useful to do full-sky imaging. In the case of MWA, the tile beam can have strong sidelobes at a distance of more than 90\degree from the pointing centre. Imaging these sidelobes might be relevant because it might either be scientifically interesting (e.g. when searching for transients), or be useful for self-calibration or full-sky deconvolution. For example, the latter is useful when a strong source in a sidelobe affects the sensitivity near the pointing centre.

To image the full sky, an observation is split into short snapshots, their phase centre is changed to zenith and subsequently a suitable area is imaged. Depending on desired resolution and dynamic range, this might require very large images. To make an image at the MWA resolution, the image needs to have approximately 10K pixels in width and height. The $w$-values will be very small at zenith, because only the differential heights of the antennas will contribute to $w$-value at zenith. For the MWA, the maximum height difference between tiles is $8.5$m. The MWA tiles are in fact on a slight slope, and by fitting a plane to the antennas and changing the phase centre towards the normal of that plane, the maximum $w$-value decreases to $5.5\textrm{m} / \lambda$. In the following sections, when discussing MWA zenith direction we will refer to this optimal $w$-direction.

\subsection{W-snapshots} \label{sec:snapshot-imaging}
\begin{figure*}
\begin{center}
\includegraphics[width=8cm]{img/vela-normal-projection}
\includegraphics[width=8cm]{img/vela-zenith-projection}
\caption{13-min MWA observation of supernova remnants Vela and Puppis A. Left: normal projection centred on Puppis A, right: phase-rotated to zenith to reduce $w$-values, shifted towards Puppis A during imaging. Both Stokes-I images have been made with WSClean using the Cotton-Schwab clean algorithm. No beam correction was applied. TODO time difference and more discussion, replace with deeper images.}
\label{fig:vela-projection-example}
\end{center}
\end{figure*}
To get optimal imaging performance, the $w$-values can be kept small by phase-changing towards zenith. As shown, all-sky imaging can be quite efficient with this technique. However, if one is not interested in imaging the whole sky, and the direction of interest is far from zenith, the computational overhead of making all-sky images is undesirable.

Instead of imaging the whole sky, the image can be recentred from $(l,m)$ to $(\hat{l},\hat{m})=(l+\Delta l,m+\Delta m)$ by performing the substitution $(l,m)\leftarrow(\hat{l},\hat{m})$ in Eq.~\eqref{eq:wstacking}:
%When the Fourier transform is performed, the $w$ term has been corrected (either by w-stacking or by w-projection). Therefore, if $U(u, v)$ is the input to the 2-dimensional Fourier transform, the transform is described by
\begin{align}\notag
\frac{I'(\hat{l}, \hat{m})\left(w_e - w_0\right)}{\sqrt{1-\hat{l}^2-\hat{m}^2}} = \int\limits_{w_0}^{w_e} e^{2\pi i w(\sqrt{1-\hat{l}^2-\hat{m}^2}-1)} \cdot \\
\iint e^{2 \pi i \left(u\Delta l + v\Delta m\right)} V(u,v,w) e^{2\pi i \left(ul + vm\right)} du dv dw.
\end{align}
In words, recentering an image involves accounting for the position shift during the $w$-correction and final scaling, and shifting the visibilities in phase by multiplication with $e^{2 \pi i \left(u\Delta l + v\Delta m\right)}$ prior to the imaging. By doing the inverse corrections during visibility prediction, such a shifted visibility set can be cleaned with major iterations similar to a non-shifted set.

A recentred image is in the projection of the zenith tangent plane, and will need an additional regridding step before multiple snapshots can be added together. A recentred image can be stored in a Flexible Image Transport System (FITS) file, which allows setting the centre of tangent projection with the \texttt{CRPIXi} keyword \citep{wcs-in-fits}. Common viewers such as Kvis and DS9 support such FITS files and display their coordinates correctly. An example of the difference in projection is given in Fig.~\ref{fig:vela-projection-example}, which displays a 13~min MWA observation imaged with WSClean. Cleaning an image that is in a different projection yields slightly different results, but qualitativily the two images are clearly of equal accuracy.

@@

\section{Analysis} \label{sec:analysis}
We will now analyse the performance and accuracy of our implementation, and compare it with the w-projection implementation in CASA. For the analysis, we use imaging parameters which are common for MWA Imaging. When a specific parameter value is not mentioned, the settings from Table~\ref{tbl:default-parameters} are used. The number of $w$-projection planes in w-projection is kept equal to the number of $w$-layers in w-stacking, and is set to the right hand side of Eq.~\eqref{eq:nwlayers-bound}. This yields 97 $w$-projection planes or $w$-layers at 10\degree zenith angle. Some configurations with high $w$-values fail to image with CASA, presumably because the $w$-kernels become to large.

TODO discuss CASA version

TODO discuss desktop computer specs
\begin{table}%
\caption{Parameter values used during benchmarks, unless otherwise mentioned.} \label{tbl:default-parameters}%
\begin{center}\begin{tabular}{rl}%
\hline
Array & MWA 128T \\
Image size & $3072 \times 3072$ \\
Angular resolution & $0.72'$ \\
Number of visibilities & $3.5 \times 10^8$ \\
Time resolution & 2~s \\
Frequency resolution & 40~kHz \\
Observation duration & 112 s\\
Bandwidth & 30.72 MHz (768 channels)\\
Central frequency & 182 MHz \\
Zenith angle at phase centre & 10\degree \\
Max $w$-value for phase centre & 172 $\lambda$ (283 m) \\
Nr. polarizations in set & 4 \\
Imaged polarization & XX \\
Imaging mode & multi-frequency synthesis \\
\hline
\end{tabular}\end{center}\end{table}

\subsection{Performance}
\begin{figure}
\begin{center}
\includegraphics[width=8cm]{img/benchmark-nsamples/nsamples}
\caption{Performance dependency on the number of imaged visibilities for imaging at zenith. Both methods are linearly dependent on the number of visibilities. WSClean is 2-3 times faster than CASA. Error bars show 10$\sigma$ level.}
\label{fig:timing-nsamples}
\end{center}
\end{figure}

In Fig.~\ref{fig:timing-nsamples} shows the time it takes to image a 2-min MWA snapshot at zenith with varying time integration. Time integration affects the number of visibilities to be gridded without changing the maximum $w$-value. Both methods have a linear time dependency on the number of visibilities. Because of the small $w$-values at zenith, the two methods are expected to be almost identical. However, the two methods show a factor of 2-3 difference, which might be attributed to different choices in optimizations. The linear dependency implies that gridding is dominating the cost.

\begin{figure}
\begin{center}
\includegraphics[width=8cm]{img/benchmark-zenith-angle/za}
\caption{Effect of zenith angle (ZA) of the phase centre on performance. Image width and height are $2048$ for the dashed line and $3072$ for the solid line. For the latter, CASA fails to image observations with ZA $\ge 15$. ZA dramatically affects the computational cost of w-projection, and WSClean is more than an order of magnitude faster at ZA $\ge$ 10\degree with a $3072$ image size (or ZA $\ge$ 30\degree at $1536$.) }
\label{fig:timing-zenith-angle}
\end{center}
\end{figure}

In Fig.~\ref{fig:timing-zenith-angle}, the computational cost as function of the zenith angle is plotted.

\begin{figure}
\begin{center}
\includegraphics[width=8cm]{img/benchmark-channels/channels}
\caption{...}
\label{fig:timing-channels}
\end{center}
\end{figure}

\subsection{Accuracy}
TODO show aliasing errors for too few wlayers.
\subsection{Gridding} \label{sec:gridding}
During gridding of the visibilities in $uv$-space, it is useful to apply a gridding convolution kernel. A common kernel function is a windowed sinc function. A low-pass filter decreases the flux of sources outside the field of view, and thus helps to attenuate aliased ghost sources and sidelobes (see \citealt{post-correlation-filtering}) that arise because of the assumed periodicity during the inverse FFT. By supersampling, a convolution kernel also makes it possible to place samples more accurately at their $uv$ position, thereby lowering decorrelation.

@@

@@

\section{Application of WSClean to the MWA}
The MWA correlator splits observations in data products of typically a few minutes
TODO

TODO: calculate when to use w-snapshots
